---
title: "What makes a house valuable?"
author: "Outstanding student 1, Awesome student 2 and Great student 3"
date: "31/11/16"
subtitle: A reproducible analysis for the Boston housing data
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
# include this code chunk as-is to set options
knitr::opts_chunk$set(comment=NA, prompt=TRUE)
library(Rcmdr)
library(car)
library(RcmdrMisc)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
# include this code chunk as-is to enable 3D graphs
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```

This short report shows a simple and non-exhaustive analysis for the price of the houses in the `Boston` dataset. The purpose is to quantify, by means of a multiple linear model, the effect of 14 variables in the price of a house in the suburbs of Boston. 

We start by importing the data into `R` and considering a multiple linear regression of `medv` (median house value) in the rest of variables:
```{r}
# Import data
library(MASS)
data(Boston)
```

```{r}
mod <- lm(medv ~ ., data = Boston)
summary(mod)
```
The variables `indus` and `age` are non-significant in this model. Also, although the adjusted R-squared is high, there seems to be a*clear non-linearity:
```{r}
plot(mod, 1)
```

In order to bypass the non-linearity, we are going to consider the non-linear transformations given in Harrison and Rubinfeld (1978) for both the response and the predictors:
```{r}
modTransf <- lm(I(log(medv * 1000)) ~ I(rm^2) + age + log(dis) +
                  log(rad) + tax + ptratio + I(black / 1000) +
                  I(log(lstat / 100)) + crim + zn + indus + chas +
                  I((10*nox)^2), data = Boston)
summary(modTransf)
```
The adjusted R-squared is now higher and, what is more important, the non-linearity now is more subtle (it is still not linear but closer than before):
```{r}
plot(modTransf, 1)
```

However, `modTransf` has more non-significant variables. Let's see if we can improve over the previous model by removing some of the non-significant variables. To that aim, we look for the best model in terms of the Bayesian Information Criterion (BIC) by `stepwise`:
```{r}
modTransfBIC <- stepwise(modTransf, trace = 0)
summary(modTransfBIC)
```
The resulting model has a slightly higher adjusted R-squared than `modTransf` with all the variables significant.

We explore the most significant variables to see if the model can be reduced drastically in complexity.
```{r}
mod3D <- lm(I(log(medv * 1000)) ~ I(log(lstat / 100)) + crim, data = Boston)
summary(mod3D)
```

It turns out that **with only 2 variables, we explain the 72% of variability**. Compared with the 80% with 10 variables, it is an important improvement in terms of simplicity: the logarithm of `lstat` (percent of lower status of the population) and `crim` (crime rate) alone explain the 72% of the variability in the house prices. We add these variables to the dataset, so we can call `scatterplotMatrix` and `scatter3d` through `R Commander`,
```{r}
Boston$logMedv <- log(Boston$medv * 1000)
Boston$logLstat <- log(Boston$lstat / 100)
```
and conclude with the visualization of: 

1. the pair-by-pair relations of the response and the two predictors;
2. the full relation between the response and the two predictors.
```{r}
# 1
scatterplotMatrix(~ crim + logLstat + logMedv, reg.line = lm, smooth = FALSE,
                  spread = FALSE, span = 0.5, ellipse = FALSE,
                  levels = c(.5, .9), id.n = 0, diagonal = 'histogram',
                  data = Boston)
```
```{r webgl = TRUE}
# 2
scatter3d(logMedv ~ crim + logLstat, data = Boston, fit = "linear",
          residuals = TRUE, bg = "white", axis.scales = TRUE, grid = TRUE,
          ellipsoid = FALSE)
```
